---
title: "Cyclistic Case Study Analysis & Visualization through R"
author: "Palash Jana"
date: "2025-07-14"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Source

I downloaded and reviewed Cyclistic’s historical trip data, which is publicly available via [Divvy 2020 Q1](https://docs.google.com/spreadsheets/d/179QVLO_yu5BJEKFVZShsKag74ZaUYIF6FevLYzs3hRc/template/preview#gid=640449855). The data consists of 1 month (Jan-2020) of trip history files in .xlsx format. It contains information such as ride ID, start and end time, start and end station, user type (casual vs. member), bike type, and ride duration.The data is organized in a tabular format, and each row represents a unique bike trip. The most critical column for my analysis is member_casual, which distinguishes between members and casual riders.

## Process Phase (through •	Google Sheets)

I began processing the Cyclistic trip data from the last 1 month to prepare it for analysis. I used Google Sheets for data processing because it allows for cloud storage, collaboration, and basic formula-based data transformation — and it works well within the memory limits.

#### Tools Chosen:
  * Google Sheets (for cleaning and transformations)
  * Google Drive (for organizing and storing data)

#### Key Steps Taken:

1.	Downloaded the .xlsx files.
2.	Created folders on my Google Drive:
    * One for original .xlsx files
    * One for converted Google Sheets
3.	Uploaded the .xlsx into Google Sheets and saved a copy in the appropriate subfolder.
4.	For the dataset:
    * Added a column ride_length using the formula =ended_at - started_at
      (in HH:MM:SS format using Format > Number > Time)
    * Added a column day_of_week using the formula =WEEKDAY(started_at,1)
      (where 1 = Sunday, 7 = Saturday)
    * Verified time zone consistency across datasets.
    * Checked for:
    
          	Missing values (e.g. station names or user type)

          	Negative ride lengths or zero duration

          	Duplicated or incomplete rows

### Cleaning & Transformation Summary:

* Removed rows where ride_length was negative or zero.
* Standardized column names.
* Filtered out rows with missing station data.
* Ensured that only valid ride types and user types (member/casual) remained
* Verified formatting consistency in rideable_type and member_casual columns.

### Data Integrity:

* The file was successfully processed with the same structure and format
* Confirmed that no PII (personally identifiable information) is present
* Retained only relevant fields: start time, end time, station info, ride type, user type, etc.

### Set CRAN mirror for non-interactive sessions

```{r}
options(repos = c(CRAN = "https://cloud.r-project.org"))
```



### Install Required Libraries


install.packages("readxl")
install.packages("tidyverse")
install.packages("lubridate")
install.packages("ggplot2")


### Load Required Libraries

```{r Load Libraries}
library(readxl)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(dplyr)
```

### Import the Dataset

```{r}
rides <- read_excel("C:\\Users\\user\\Desktop\\Data Analyst Projects\\Google Data Analytics Course\\8_Google Data Analytics Capstone_Complete a Case Study\\Capstone Project\\cyclistic_jan_2020.xlsx", sheet = 1)
```


### Ensure datetime columns are in proper format

```{r Ensure datetime columns are in proper format}
rides <- rides %>%
  mutate(
    started_at = ymd_hms(started_at),
    ended_at = ymd_hms(ended_at),
    time_of_day = hour(started_at),
  )
```

### Calculate Average Ride Length by User Type
```{r Calculate Average Ride Length}
avg_ride_length <- rides %>%
  group_by(member_casual) %>%
  summarize(avg_ride_length = mean(ride_length))
```

### To convert date and time to only minutes.
```{r convert date and time to only minutes}
avg_ride_length$ride_length_mins <- as.numeric(lubridate::hms(substr(avg_ride_length$avg_ride_length, 12, 19))) / 60
print(avg_ride_length)
```


### Count Rides per Day of the Week

```{r Count Rides per Day of the Week}
rides_per_day <- rides %>%
  group_by(member_casual, day_of_week) %>%
  summarize(total_rides = n()) %>%
  arrange(member_casual, day_of_week)

print(rides_per_day)
```

### Summarize Usage by Time of Day

```{r Summarize Usage by Time of Day}
usage_by_hour <- rides %>%
  group_by(member_casual, time_of_day) %>%
  summarize(total_rides = n())

print(usage_by_hour)
```

## Visualization through R

### Column chart on Average Ride Length by User Type

```{r Column chart on Average Ride Length by User Type}
ggplot(data = avg_ride_length) +
  geom_col(mapping = aes(x = member_casual, y = ride_length_mins, fill = member_casual)) +
  labs(
    title = "Average Ride Length by User Type",
    x = "User Type",
    y = "Average Ride Length (minutes)"
  )
```


### Column chart on Total Rides per Day by User Type

```{r Column chart on Total Rides per Day by User Type}
ggplot(data = rides_per_day) + 
  geom_col( mapping = aes(x = day_of_week, y = total_rides, fill = member_casual)) +
  labs(
    title = "Total Rides per Day of the Week by User Type",
    x = "Day of the Week",
    y = "Number of Rides"
  )
```



### Line chart on Ride Usage by Hour of Day

```{r Line chart on Ride Usage by Hour of Day}
ggplot(usage_by_hour, aes(x = time_of_day, y = total_rides, color = member_casual, group = member_casual)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "Ride Usage by Hour of Day",
    x = "Hour of Day (0–23)",
    y = "Total Number of Rides",
    color = "User Type"
  ) +
  scale_x_continuous(breaks = 0:23) +
  theme_minimal()
```












